% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\author{}
\date{}

\begin{document}

\hypertarget{romans-1-esv}{%
\section{Romans 1 (ESV)}\label{romans-1-esv}}

\textbf{Greeting} \emph{1 Paul, a servant of Christ Jesus, called to be
an apostle, set apart for the gospel of God, 2 which he promised
beforehand through his prophets in the holy Scriptures, 3 concerning his
Son, who was descended from David according to the flesh 4 and was
declared to be the Son of God in power according to the Spirit of
holiness by his resurrection from the dead, Jesus Christ our Lord, 5
through whom we have received grace and apostleship to bring about the
obedience of faith for the sake of his name among all the nations, 6
including you who are called to belong to Jesus Christ,} \emph{7 To all
those in Rome who are loved by God and called to be saints:} \emph{Grace
to you and peace from God our Father and the Lord Jesus Christ.}

The gospel of God is clearly explained in here and consists of (1) the
prophecy, (2) Jesus human and divinity nature, (3) His death and (4) His
resurrection. However, it is possible to summarize and organize this
text even further. * What is the Gospel? It is these four points
mentioned above. * What is the goal of the Gospel? To bring about the
obedience of faith (v.50). * Why is there a Gospel? For the sake of His
Name among all the nations. * For whom is the Gospel? This is not
answered here---although we know it is meant for everyone because God
did not limit us to exclude certain people, although He is the One Who
decides who will hear it---but it includes those who are in Jesus Christ
(v.6). So even Christians need to hear the Gospel \emph{again}.

\textbf{Longing to Go to Rome} \emph{8 First, I thank my God through
Jesus Christ for all of you, because your faith is proclaimed in all the
world. 9 For God is my witness, whom I serve with my spirit in the
gospel of his Son, that without ceasing I mention you 10 always in my
prayers, asking that somehow by God's will I may now at last succeed in
coming to you.}

The notes on verses 11-13 may indicate as if Paul was not doing any of
his out of love, but purely out of obedience to God. So perhaps there
was love involved for God, but not for man. However, these verses
dispute this claim as Paul indicates that God is his witness of that he
\emph{always} mentions the addressees (i.e.~the Romans) in his prayers.
This in itself is proof enough. Could he be praying out of obedience?
Yes, but it would be far-fetched to think that Paul prayed because he
had to and included in his prayer those he had to pray for. Even though
prayer is an ``obligatory'' part for believers---any believer prays, if
he does not, he cannot be deemed as one---the contents of these prayers
are to be freely filled in by the prayer. So Paul could have prayed for
anything he wanted, such as the saving of the city of Jerusalem, his
fellow Jews in the diaspora or his own wellbeing, but he chose to
include them \emph{always}, which to me sounds much more than merely a
mention of them in his prayers ``every day''.

\emph{11 For I long to see you, that I may impart to you some spiritual
gift to strengthen you--- 12 that is, that we may be mutually encouraged
by each other's faith, both yours and mine. 13 I do not want you to be
unaware, brothers, that I have often intended to come to you (but thus
far have been prevented), in order that I may reap some harvest among
you as well as among the rest of the Gentiles.}

Paul is clear here that he comes to get results. Our work for the
Kingdom of God is not intended to be for the purpose of leisure. He says
in verse 12 that he \emph{longed} to see them, but then he adds the
clause ``that I may impart to you\ldots{}''. Would that mean that he
does not truly long to see them? Is that something like loving someone
\ldots{} because she is beautiful? The latter clause would diminish the
purity of love, it seems, just like it seems to diminish Paul's longing.
He later explains that strengthening them is not the only reason, Paul
Himself wants to be encouraged as well. This actually only adds to
negative feeling of \#todo/opzoeken/engels niet oprechte gevoelens of
bedoelingen So Paul comes for a purpose, and he does it so that he can
reap the benefits of it himself as well, and he is looking for results.
Where is the love here, one might think? It is true, Paul's first and
foremost job is to obey God and not act out of whatever or whomever he
loves. But this does not mean that both things have to be exclusive.

\emph{14 I am under obligation both to Greeks and to barbarians, both to
the wise and to the foolish. 15 So I am eager to preach the gospel to
you also who are in Rome.}

In a certain sense it seems like Paul is referring to those in Rome as
the foolish, because he uses the word ``so'' which seems to refer to the
latter used description, which is ``foolish''. Of course another
interpretation is that ``so'' refers to ``the wise and the foolish'' and
thus anything in between as well. But even if the addressees are
considered the foolish, would that be wrong? It would be foolish
compared to the Greek. Now that would not be harsh for two reasons.
First of all the Greeks were very intelligent people to whom many
literary, philosophical, methodological and mathematical works may be
attributed. Almost any other people would be like barbarians when
compared with them. Second of all, however, the Greeks were barbaric
\emph{morally speaking} compared to the Jews and Christians. So if
Christians were ``barbaric'' compared to Greeks, then that could only be
a compliment. Being anything other than a Greek on the field of religion
would always be a compliment.

\textbf{The Righteous Shall Live by Faith} \emph{16 For I am not ashamed
of the gospel,}

Why would Paul mention anything here about shame? Why would it be
shameful to have a God Who was foretold, is simultaneously human, and
who died and rose again? It is exactly for those reasons. A god who is
human is a lesser god---a \emph{hero} or a demi-god, as the Greeks would
call---lower than the other real gods, unless they are made into gods. A
god who dies has weaknesses and is not almighty. One who rose
again---out of the Hades that is---indicates that he was weak enough to
be thrown in it in the first place. In other words, compared to the
Greek and Roman gods, and perhaps the Babylonian and all other pagan
gods, Jesus Christ was less than a god. They do not understand many
things about our Lord, however. For one, they do not understand the
concept of the Trinity, and of a Person Who can be god and human at the
same time. A hundred percent plus a hundred percent adds up to two
hundred percent according to logic, but they do not realize that that is
human logic. Once they do understand, they would realize that a human
god would only add to the godliness of Jesus Christ. Jesus felt all the
pain as a human did, a weak one at that, unlike Simson who must have
felt pain when fighting the Philistines or Hercules when he had this
poisoned mantle covering him, but who were very strong to begin with.
Jesus had a normal body, not of a soldier, perhaps a bit stronger than
average due to carpenting, but nothing out of the extraordinary. He had
feelings, much \emph{more} than human beings exactly \emph{because} of
His godliness. As God He could feel everything we were feeling and He
did so on the cross, where He suffered so much that He instantly died,
of all the feelings and pain of all people before and after Him.

So the first argument that Jesus was only a demi-god is not true,
because the Bible uses the wholly different concept of the God-head
(three in one) which has never appeared throughout history in any
religion or literature.

The second argument of Jesus' weakness in His humanness
\#todo/opzoeken/engels is invalid because His humanness actually makes
Him stronger than other gods.

Now a third argument people use is His death. Which is also one of the
main reasons why the Jews did not believe Jesus to be the Messiah. Death
shows weakness, because apparently one was not strong enough to defend
Himself. It also shows terminality, \#todo/opzoeken/engels een
eindigheid, because God's reign would be forever, although (some) Jews
like Ben Shapiro \#tags/famouspeople believe the Messiah is not a Person
who will reign forever, but just a human, a political figure who will
accomplish things on Earth. They are forgetting one of the most
important traits of a real God, however, which other gods do not have,
which is meekness. I rarely hear about meekness or self-sacrifice in
Buddhism, Hinduism and Islam. Yes, there is the story where Buddha
sacrificed himself by killing one person (and perhaps getting executed
for that as well) because he knew this person would kill a hundred.
\#todo/nogaftemaken

\emph{for it is the power of God for salvation to everyone who believes,
to the Jew first and also to the Greek. 17 For in it the righteousness
of God is revealed from faith for faith, as it is written, ``The
righteous shall live by faith.''}

This is a difficult sentence. It says that God's righteousness is ``from
faith for faith''. What would that mean? God's righteousness must be
\emph{revealed}, so not everyone can see it or know about it unless God
chooses to reveal it to someone. Now it is revealed from faith. So faith
is like a pair a of glasses with which righteousness can be seen. What
many do not know, is that God's righteousness contains an aspect of
time. As we have seen in both the Pharaoh in Moses' time, and Noah's and
Jonah's stories is that people were rewarded or punished in a context of
time. God did not punish people directly, but He had his time. Now
delayed punishment requires patience, when one wants justice to occur
immediately, but patience is built on a greater thing called reasoning.
Why would anyone wait with punishment? Because proof of one's innocence
could appear or remorse or retribution ??. If a person on death row can
provide more information about the criminal network he used to work for,
would that not be worth anything? If he converts to Christianity, would
he then not be valuable as a mentor to many other potential criminals?
Waiting can have a positive effect on the outcome. One needs reasoning
for that. But what if one cannot see any changes? What if none of the
above applies to this particular criminal? Then only one thing remains:
\emph{faith}. Through faith one can have hope, hope that change is
\emph{possible}.\\
There are so many stories where the whole world did not see any hope for
a particular person, yet this person arose from the surface all because
one person still believed in him. Take for example Ben Carson's story
depicted in the movie \emph{Gifted Hands} starring Cuba Gooding Jr.~,
or---perhaps---the movie \emph{Jerry McGuire} where Tom Cruise believes
in Cuba Gooding Jr.~ If we believe that God has a plan with
\emph{delaying} His judgment, we may be able to see His righteousness.
In other words, it is only through belief and for those who believe.
\#church/material

\textbf{God's Wrath on Unrighteousness} \emph{18 For the wrath of God is
revealed from heaven against all ungodliness and unrighteousness of men,
who by their unrighteousness suppress the truth.}

The wrath of God too is revealed. Though it does not say here, it is
revealed to those God wants it to be revealed. How is ungodliness
punished? Not everyone can see this. Again it is through faith only that
we can see this. We see some things happening on Earth and we do not
know how to interpret this. We see for example tumult in Iran and chaos
in Myanmar. We see innocents dying by the thousands. The question arises
in us then where righteousness is. Why are innocents being punished? The
real question is, however, are they really innocent? It requires a
revelation from God to see the answer. Did not they and their
forefathers participate in hundreds if not thousands of years of pagan
worship? Did they not persecute Christians directly or indirectly by
assisting their government or voting for policies against Christians?
Myanmar's current (female) leader, \emph{Aung Kie\ldots{}}
\#todo/opzoeken was chosen democratically. Yet she kills muslims by the
thousands. Are the people not responsible for that either? The same goes
for almost any dictator. They have all been chosen by the people at some
point. To name a few more, Che Guevara, Fidel Castro, Mao Ze Dong and
Hitler \#tags/ideology/communism. Even when some of those ``innocent''
people did not actively persecute Christians, who knows whether they
might have done so, had they been in the right situation and time? Only
God knows and only He can be the judge of that.

What is furthermore written in this verse about the sin of man, is that
man suppresses the truth. That alone is worthy of severe punishment. It
is the covering up of the truth that causes all trouble. Can anyone name
one example of a problem that would have been solved if everyone was
truthful? If a murder intents to kill anyone and he is truthful about
it, he might be stopped in his actions even before they happen. If he
has already done it, he will be apprehended to prevent him from doing it
another time. Of course, within marriages the truth is not enough to
solve a problem, even when one commits adultery and immediately admits
it. Remorse and willingness need to be present as well. But generally
speaking, solving a problem always requires the truth. The question we
need to ask is always ``What caused this problem?'', and the
\emph{cause} is the truth.

\emph{19 For what can be known about God is plain to them, because God
has shown it to them. 20 For his invisible attributes, namely, his
eternal power and divine nature, have been clearly perceived, ever since
the creation of the world, in the things that have been made. So they
are without excuse.}

Now one might think that my statement about suppressing the truth is a
bit harsh, but as it says here, the are without excuse. Do I think
adults are deliberately suppressing the truth to their children about
God, His creation and His Almight? I do think that many atheists and
heathens do not know God, but I do not believe that they have no doubts
at all about their beliefs. Everyone, no matter how little, knows or
feels that atheism is empty and hopeless, that evolution is
counterintuitive---and still they teach their children something that is
against their own conscience. All Leftists know that the gender-alphabet
starting with LGBT is going too far and that for such a small percentage
of people systems need to change, for reasons that are not even
scientifically proven (namely that gender is something between the ears)
and are counterintuitive, namely that children know they are of the
opposite sex at age 3. People from other religions, such as Islam,
Buddhism and Hinduism, also know that something is wrong with their
religion. How can it be that some people are worth more than others?
Even including the concept of reincarnation does not resolve this
problem, because human beings can still feel that beggars, even if they
deserve to be one because of their previous life, should be treated
fairly, simple \emph{because they are human beings}. No religion or
philosophy can talk its way out of that. It is in us because of our
intrinsic worth in God's eyes.

\emph{21 For although they knew God, they did not honor him as God or
give thanks to him, but they became futile in their thinking, and their
foolish hearts were darkened. 22 Claiming to be wise, they became fools,
23 and exchanged the glory of the immortal God for images resembling
mortal man and birds and animals and creeping things.}

I am not sure how much the pagans knew God, neither do I know how much
the Jewish people of the first century knew Him. Except for Jesus'
miracles I do not think either of them have seen much of God. The Jewish
people have their ancestors and their Bible and both have creation as
the proof of God's existence. We can see here a logical order in which
they replaced God with all kinds of attributes. The first thing is
denying God. They knew God---whatever that means, it at least indicates
that they knew Who God was and that they should honor and thank
Him---but it is exactly those things they purposely neglect to do. As
with many or most things in life, there are multiple ways of going, but
when one does not go the right way, he has to \#todo/opzoeken/engels
zichzelf wringen in allerlei bochten he has to go through all kinds of
loops (or hoops \#todo/opzoeken/engels?) just to be able to explain his
choices. This is true for the truth and logic. When a murderer claims he
has not killed someone, he must create an alibi, which could be a
person, and this person as well needs to have an alibi that he saw him,
and so forth, which is not possible anymore after a few steps. When the
evolutions say that the world is created in millions of years they need
many ice ages to explain the formation of our current world, while
Occam's Razor can be applied here as well, it is the simplest
solution---one big ice age after Noah's flood---which can explain the
world's formation as well. Now at a certain moment this becomes flat out
lying to them. They were lying in the beginning already, but at least
they were mixing in truth with lies. No they are flat out lying because
they do not have any other choice. This is what we see with the
Leftists. In the beginning they could mix in some lies with the truth,
the lies being that they want to be good to black people and other
minorities, they want to do good to the people, but after so many things
are exposed about their politics, that this is just a modern
\#todo/opzoeken/engels modern jasje, in a new jacket? of socialism as it
was implemented in the previous century---which did not work by the way
and will never work---people like Samantha Nixon (I do not know if she
was first) simply admitted: ``Ok, if this is what you want to call it,
then we'll admit. We want socialism. But it is not standard socialism,
but \emph{democratic socialism}.''

::\emph{24 Therefore God gave them up in the lusts of their hearts to
impurity, to the dishonoring of their bodies among themselves, 25
because they exchanged the truth about God for a lie and worshiped and
served the creature rather than the Creator, who is blessed forever!
Amen.}:: \#biblestudy/memorization/todo

Here too we find a major principle throughout human lives and which can
be found within sin. It is that God gives up people to their own
longings and that people rather worship the created than the creator,
regardless whether this concerns God or not. Throughout the New
Testament we see that Jesus does not force His will or world views on
others. He attracts people and persuades them with reasoning. If they
want to they can follow Him and if they do not, they can do the thing
they want themselves. The same it is with God. God intervenes, as He has
done with the Ninevites, but that is because He knew that in their
hearts that generation was willing to give up their lust and impurity.
But if a people is not ready to do so, then God leaves this person,
unless He is graceful and tries again. I am not sure of the latter, that
is, whether God tries multiple times to persuade peoples, but from what
I see in the Old Testament God did not kill the Hittites, Peruzzites
\#todo/opzoeken etc. in one time, but gave them multiple chances to give
up the fight. We do not know how much knowledge or opportunities these
pagan peoples in and around Canaan were given about our Lord Almighty,
but we know that He is righteous.

\emph{26 For this reason God gave them up to dishonorable passions. For
their women exchanged natural relations for those that are contrary to
nature; 27 and the men likewise gave up natural relations with women and
were consumed with passion for one another, men committing shameless
acts with men and receiving in themselves the due penalty for their
error.}

In verse 25 we read that God gave them up to their lusts and here in
verse 26 God gives them up to their passions. Also, the people
dishonored their bodies by exchanging the truth for lies, and exchanging
natural relations for unnatural ones, respectively. It is this specific
order in which things happen. Granted, homosexuality has always existed,
long before there were ``scientists'' claiming that gender is something
between the ears, but it was always common knowledge that this was not
something natural. So now, in these days, the Left exchanges the truth
for the lie, and many Christians, who should be on the Right, slowly
(but still too fast) start leaning that way and giving in. And as soon
as that happens the action starts as well. Lies cannot stand the truth,
just like darkness hates light. The Bible is the truth, so if one wants
to follow the lie, he needs to cut out the Bible from His life. So now
some churches such as the \emph{Alexanderkerk} in Rotterdam accept gay
marriages, just like the Remonstrantse Kerk in the Netherlands. The
churches are now Law abiding ``except for one rule''. No, if one starts
cutting in the Bible, or in the truth in that case, the whole truth will
disappear. Once again, as verse 25 and 26 say, if one persists in
exchanging the truth for a lie, God will give them up to their lusts. In
due time these churches will avail themselves.

\emph{28 And since they did not see fit to acknowledge God, God gave
them up to a debased mind to do what ought not to be done.}

So after they have been given up to their own lust and passions, God
gave them up to their debased mind as well. Now this seems all the more
true of the Left. The Left is using their mind in reaching their goals.
Their strategies are well thought of. They have purposefully pointed the
media, education and Hollywood in the same direction and thought of the
strategy to mark who does not agree with them as bigots, homophobes or
xenophobes. It requires a lot of planning to wag the dog in this way.
The sad thing, however, is as verse 28 says: their mind is debased. They
have arrived at the stage where they do not even know what or that they
are doing wrong anymore. Those influenced by the Left have never known
perhaps. But the master strategists, those higher ``ranked'' or longer
seating than Hillary Clinton, they used to know better, but not so
anymore. They minds have become debased and their sin is all the more
increasing. It is different from a person born with a debased mind or
intelligence and who really does not know any better. These persons have
hardened their hearts and their logical thinking and God, Who has now
hardened their hearts even more, will punish them for all their sins,
but also---just like in the case of Pharaoh---use these atrocities for
the good of His people.

\emph{29 They were filled with all manner of unrighteousness, evil,
covetousness, malice. They are full of envy, murder, strife, deceit,
maliciousness. They are gossips, 30 slanderers, haters of God, insolent,
haughty, boastful, inventors of evil, disobedient to parents, 31
foolish, faithless, heartless, ruthless.}

What debased means in the previous verse is more explicitly explained
here. The sins range here from very obvious sins to less obscure sins.
We see that these sins which included only the worshipping of other
gods, and sinning towards God, now includes sinning towards people as
well---though that is not very clear from the text. What is clear,
however, is that there sin is ever increasing. The type of sin, the
amount of sins and the way their bodies are harmed, and the continuation
of all of these. Being boastful seems like a small sin, but remember
that their boastfulness is based on their pride of knowledge. They
\emph{think} they are wise and influence other people to do so as well.
So even if they are not hurting other people, they are causing them to
sin when these people are following them.

\emph{32 Though they know God's righteous decree that those who practice
such things deserve to die, they not only do them but give approval to
those who practice them.}

Just when you think that these people Paul is speaking of are the worst
of their kind, it appears there is something that makes them even worse.
So aside from sinning against God, the conscience, their mind, their
body and other people, they now stimulate other people to sin as well.
Let us look at the former verse and this one together and see how much
this resembles the Left party in the United States. The Left hates God
and everything He commands and stands for. They are in fact inventors of
evil, by having invented a multitude of devices, institutions and
policies which discriminate and prey upon the weakest people in society,
such as \emph{Planned Parenthood} which preys on the weakness of
(especially young) women of other ethnicity who have trouble being a
single mother; \emph{Food stamps} have been created to help the weakest
in society, but the excessive distribution of these keeps the poor poor
without any incentive for labor; They have also invented things such as
the decoupling of money with the gold standard and perhaps they have
invented something like \emph{Federal Reserve Bank} which controls all
money to keep America poor. In this verse we see that the Left which has
done all of the above, endorses people to go further than they have and
to accomplish their dream of open borders for America, free education,
medical care---even for illegal immigrants---repayment of all student
debts, and what not more. All of this already depletes what is left of
federal reserves of the United States

\#biblestudy/devotionals/romans \#tags/politics \#tags/lgbt

\end{document}
